```{r set}
#| include: false 
library(tidyverse) 
library(cmdstanr)
library(bayesplot)
```

# Model

$$\begin{align}
log(agb) \sim \mathcal N (log(\mu), \sigma) \\
\mu = \sum_{i \in p} \gamma_s \times 0.0673 \times [wd_i \times dbh_i^2 \times \frac{\alpha_f \times dbh_i}{\beta_f+dbh_i}]^{0.976} \\
\gamma_s = exp(\epsilon_s \times \sigma_s) \\
\alpha_f \sim \mathcal N (\alpha_0,\sigma_a) \\
\beta_f \sim \mathcal N (\beta_0,\sigma_b) \\
\epsilon_s  \sim \mathcal N (0,1) \\
\end{align}$$

With plot biomass $agb$ linked to the sum of tree biomass $\mu$ defined on tree wood density $wd_i$, tree diameter $dbh_i$ and tree height $h_i$ itself derived from $dbh_i$ following a Michaelis Menten law with parameters $\alpha_f$ and $\beta_f$ depending on the factor $f$ being the intersection of the $system$ and the $origin$ and a random parameters $\gamma_s$ representing species random effect.

> Modelling remark
>
> - The model is working better when modelling total AGB and not per hectare AGB that works better when computed posthoc (beware also the division by 1000 for units). **It seems also really important to keep low AGB plots for tree height scaling parameters (AGB\<10t/ha).**
> - I developed a version pre-computing log agb and mainly using a cumulative sum trick to reduce computing time (but parrallelisation is useless).

```{r data}
#| message: false
#| warning: false
trees <- read_csv2("data/all_tree_corr.csv") %>% 
  select(-1) %>% 
  mutate(system = gsub("_", " ", system)) %>% 
  select(database, id_plot, id_tree, system, origin, dbh, H, genusCorr, speciesCorr) %>% 
  rename(h = H, genus = genusCorr, species = speciesCorr) %>% 
  filter(!is.na(dbh), !is.na(h))
trees$wd <- BIOMASS::getWoodDensity(genus = trees$genus,
                                    species = trees$species,
                                    stand = trees$id_plot)$meanWD
plots <- trees %>% 
  mutate(agb = 0.0673*(wd*dbh^2*h)^0.976) %>% 
  group_by(database, system, id_plot) %>% 
  summarise(agb = sum(agb)) %>% 
  left_join(select(read_csv2("data/all_plot_corr.csv"), 
                   id_plot, area_plot.ha)) %>% 
  mutate(agb_ha = agb/10^3/area_plot.ha) %>% 
  filter(agb_ha > 0, agb_ha < 1000) %>% 
  ungroup() %>% 
  mutate(plotnum = as.numeric(as.factor(id_plot)))
trees <- trees %>% 
  left_join(plots) %>% 
  filter(!is.na(plotnum)) %>% 
  select(-agb, -agb_ha, -area_plot.ha) %>% 
  mutate(factornum = as.numeric(as.factor(paste(system, origin)))) %>% 
  mutate(speciesnum = as.numeric(as.factor(paste(genus, species)))) %>% 
  arrange(plotnum, speciesnum, factornum) %>% 
  mutate(obs = 1:n())
positions <- trees %>% 
  group_by(plotnum) %>% 
  summarise(start = min(obs), end = max(obs))
mdata <- list(
  N = nrow(trees),
  P = nrow(plots),
  S = max(trees$speciesnum),
  "F" = max(trees$factornum),
  plot = trees$plotnum,
  species = trees$speciesnum,
  factor = trees$factornum,
  agb = plots$agb,
  agb_ha = plots$agb_ha,
  area = plots$area_plot.ha,
  dbh = trees$dbh,
  wd = trees$wd,
  start = positions$start,
  end = positions$end,
  grainsize = 1
)
```

```{r fit_par}
#| eval: false
model <- cmdstan_model("models/factor_par.stan")
path <- "chains/systemorigin_par/"
if(file.exists(path))
  unlink(path, recursive = TRUE)
dir.create(path)
fit <- model$sample(
  data = mdata,
  chains = 4,
  parallel_chains = 4,
  save_warmup = FALSE,
  output_dir = path,
  refresh = 100
)
```

```{r fit}
#| eval: false
model <- cmdstan_model("models/factor.stan")
path <- "chains/systemorigin/"
if(file.exists(path))
  unlink(path, recursive = TRUE)
dir.create(path)
fit <- model$sample(
    data = mdata,
    chains = 4,
    parallel_chains = 4,
    save_warmup = FALSE,
    output_dir = path,
    refresh = 100
  )
```

```{r fitres}
#| message: false
#| warning: false
#| fig-cap: "Caption."
fit <- as_cmdstan_fit(list.files("chains/systemorigin_par/", full.names = TRUE))
fit$draws(c("lp__", "alpha", "beta", 
              "sigma", "sigma_s", "sigma_a", "sigma_b")) %>% 
  mcmc_trace() +
  theme_bw()
```

> Convergences are good. Variances could be better constrained.

```{r time}
#| message: false
#| warning: false
fit$time()$chains %>% 
  mutate(across(c("warmup", "sampling", "total"),
                seconds_to_period)) %>% 
  knitr::kable()
```

```{r pvo}
#| message: false
#| warning: false
#| fig-cap: "Caption."
fit$summary("mu") %>% 
  mutate(obs = plots$agb) %>% 
  ggplot(aes(median, obs)) +
  geom_abline(col = "darkgrey") +
  geom_segment(aes(x = median - sd, xend = median + sd), col = "darkgrey") +
  geom_segment(aes(x = q5, xend = q95), size = .1, col = "darkgrey") +
  geom_point() +
  theme_bw() +
  scale_x_log10() +
  scale_y_log10() +
  xlab("Predicted AGB") +
  ylab("Observed AGB")
```

```{r pars}
#| message: false
#| warning: false
#| fig-cap: "Caption."
fit$summary(c("alpha_f", "beta_f"), 
            quantiles = ~ posterior::quantile2(., probs = c(0.05, 0.25, 
                                                            0.5, 0.75, 0.95))) %>% 
  separate(variable, c("parameter", "f", "factornum"), convert = TRUE) %>% 
  select(-f) %>% 
  left_join(select(trees, system, origin, factornum) %>% unique()) %>% 
  select(-factornum) %>% 
  gather(quantile, value, -parameter, -system, -origin) %>% 
  mutate(system = gsub("_", " ", system)) %>% 
  mutate(metric = paste0(parameter, "_", quantile)) %>% 
  select(-parameter, -quantile) %>% 
  pivot_wider(names_from = metric, values_from = value) %>% 
  ggplot(aes(beta_q50, alpha_q50,
             col = paste0(system, "\n", origin),
             label = paste0(system, "\n", origin))) +
  geom_segment(aes(x = beta_q25, xend = beta_q75)) +
  geom_segment(aes(x = beta_q5, xend = beta_q95), size = .2) +
  geom_segment(aes(y = alpha_q25, yend = alpha_q75)) +
  geom_segment(aes(y = alpha_q5, yend = alpha_q95), size = .2) +
  geom_point() +
  ggrepel::geom_text_repel(box.padding = 0.4, ) +
  theme_bw() +
  scale_color_discrete(guide = "none") +
  ylab("Asymptotic height [ m ]") +
  xlab("Diameter at 50% height [ cm ]")
```

> Some systems are good other seems less, but this is not what the likelihood is optimizing.

```{r comp}
#| message: false
#| warning: false
#| fig-cap: "Caption."
coords <- read_csv2("data/all_plot_corr.csv") %>% 
  select(database, id_plot, long, lat)
trees$agb_chave <- BIOMASS::computeAGB(D = trees$dbh, 
                                       WD = trees$wd, 
                                       coord = left_join(trees, coords)[c("long", "lat")])
agb_chave <- trees %>% 
  group_by(database, system, id_plot, plotnum) %>% 
  summarise(chave = sum(agb_chave)) |> 
  left_join(select(read_csv2("data/all_plot_corr.csv"), 
                   id_plot, area_plot.ha)) %>% 
  mutate(chave = chave/area_plot.ha) 
plots %>% 
  left_join(agb_chave) |> 
  ungroup() %>% 
  mutate(model = fit$summary("pred_ha", "median")$median) %>% 
  gather(model, predicted, -database, -system, -id_plot,
         -plotnum, -agb, -agb_ha, -area_plot.ha) %>% 
  rename(observed = agb_ha) %>% 
  mutate(error_log = log(observed) - log(predicted)) %>% 
  ggplot(aes(system, error_log, fill = model)) +
  geom_boxplot() +
  theme_bw() +
  coord_flip()
```

> Mixed results about the advantage of the model compared to the ref. Not in the case of the forest. Very important for agroforestry.
>
> Question
>
> - Repeat with null model, system only and origin only?
> - Evaluation? Quality of fit or prediction? Cross-validation?
